---
title: Introduction to Stata
author: Benjamin Skinner
output: 
  html_document:
   toc: true
   theme: cosmo
   css: ./css/custom.css
---

```{r, echo = FALSE, message = FALSE}
require(knitr)
source('./r/helper.r')
opts_chunk$set(echo = FALSE, message = FALSE, comment = NA)
```

```{r, engine = 'bash'}
## run accompanying .do file to get log file for parsing
stata-se -b -q do ../do/lecture0_introduction.do
## convert plots used in this file to png
pdir=../plots/
glist=(h_poplt5_pr
       h_poplt5_pr_reg
       kd_poplt5_pr
       sc_poplt5_pr
       sc_poplt5_pr_1
       sc_poplt5_pr_2
       )
for i in ${glist[@]};
do
convert -density 150 -flatten $pdir$i.eps $pdir$i.png;
done
```

```{r}
## save log file in object
lf <- 'lecture0_introduction.log'
```
In this class we'll walk through some of Stata's basic
functionality. We'll also get used to the idea of interacting with
Stata through the command line and `*.do` files.

<br>

## Stata as a calculator

Stata can be used as a calculator via the `display` command. All of the
normal rules of arithmetical precedence apply to the Stata
syntax.

```{r}
start <- 'using the display command for arithmetic'
end <- 'using do files'
writeLines(logparse(lf, start = start, end = end))
```

*NB:* The `display` command can be shortened to `di`. Many
Stata commands and options are like this. The help files underline the
minimum part of the command/option that must be specified in order for
the package to understand what you want.

<br>

## Using `*.do` files

Stata syntax is stored in what's called `*.do` files. These are all of
the typed commands that you use to manipulate and analyze the data. A
properly formatted `*.do` file can be run from the command line using
the `do` command:

```{r}
start <- 'using do files'
end <- 'comment styles'
writeLines(logparse(lf, start = start, end = end))
```
One of the key skills you'll learn this year is properly annotating a
`*.do` file. Remember that these files are primarily meant to be read by
humans, and only incidentally meant to be read by computers. Stata
assumes that everything in a `*.do` file is a command unless it's preceded
by a comment sign. To set off a line of text as a comment,
place `*` or `//` in front of it. You can also use the `/* ... */` 
format for comments, which can be used on the same line as the syntax
itself:

```{r}
start <- 'comment styles'
end <- 'set working directory'
writeLines(logparse(lf, start = start, end = end))
```

<br>

## Loading Stata data files

All Stata data files are saved in the `*.dta` format. Today we'll be
using the `census.dta` file which contains information on
characteristics of the 50 states from the 1980 census.

To locate a data file, you first have to tell Stata where to look on
your computer. With some very rare exceptions, you should always use
the `cd` command to set the working directory:

```{r}
start <- 'set working directory'
end <- 'filename of dataset'
writeLines(logparse(lf, start = start, end = end))
```

*NB:* The exception to this rule would be (1) when you double-click your
 `*.do` file and have Stata configured to open automatically; and (2)
 your `*.do` file is set to work in the directory in which it is
 currently located (i.e., all the relative links are correctly
 specified).

The above directory is where I keep the class files, hence the `cd`
command doesn't really do anything. Your files will be in
a different location on your computer. Changing the working directory
just once makes it *much* easier to exchange `*.do` files across
computers. Don't place a `cd` command in your `*.do` file. This will
make collaboration much easier.

To open a Stata file, use the `use` command:

```{r}
start <- 'filename of dataset'
end <- 'taking a look at the data: list'
writeLines(logparse(lf, start = start, end = end))
```

We'll go over for other commands for importing more complex data files
later.

<br>

## Looking at the data

### `list`
We can use the `list` command to take a look at the data:

```{r}
start <- 'taking a look at the data: list'
end <- 'taking a look at the data: describe'
note <- '...and so on.'
writeLines(c(logparse(lf, start = start, end = end)[1:24],note))
```
### `describe`
As should be obvious, this usually gives too much information back. A
better place to start with a well-formatted data file is to use the
`describe` command:

```{r}
start <- 'taking a look at the data: describe'
end <- 'taking a look at the data: codebook'
writeLines(logparse(lf, start = start, end = end))
```
### `codebook`
To get more information about a single variable, the`codebook`
command is a good option:

```{r}
start <- 'taking a look at the data: codebook'
end <- 'show me the data for the first ten states'
writeLines(logparse(lf, start = start, end = end))
```

### `list` with `if` statement

If I add the condition `if _n < 11`, I can see data for only the first
ten states. `_n` represents the row number
of each observation. Since the states are in alphabetical order in the
dataset, I can use the logical statement `_n < 11` to get the first
ten:

```{r}
start <- 'show me the data for the first ten states'
end <- 'just state names and populations for the first ten states'
writeLines(logparse(lf, start = start, end = end))
```

Most of the time, I only want to see a couple of variables. In this
case, I'll use `list` with what Stata calls a `varlist` and is in fact
just a list of variables. In this case, I only choose `state` and
`pop`.

```{r}
start <- 'just state names and populations for the first ten states'
end <- 'recoding variables'
writeLines(logparse(lf, start = start, end = end))
```

<br>

#### QUICK EXERCISE
> Take a look at deaths in the first 10 states. Which is highest,
> which is lowest?

<br>

## Recoding variables

To start off with, I'm interested in knowing which states have the
largest proportion of the population under 5. The data only give the
total number of people under 5, so I'm going to need a new variable,
which will be total population under 5 divided by total population. To
create this variable I'll need Stata's `generate` command:

```{r}
start <- 'recoding variables'
end <- 'summarize the new variable'
writeLines(logparse(lf, start = start, end = end))
```

*NB:* Stata will not allow you to generate a new variable with an old
 variable's name. `generate poplt5 = poplt5 / pop` will not work
 because you already have a `poplt5` variable. This is a feature to
 make sure you don't overwrite your data accidentally.

<br>

## Summarizing data

Now that I have my new variable, let's use the summarize command to
take a look at it:

```{r}
start <- 'summarize the new variable'
end <- 'summarize with more detail'
writeLines(logparse(lf, start = start, end = end))
```

This is nice, but if I'd like even more information I should use the
`detail` subcommand, like so:

```{r}
start <- 'summarize with more detail'
end <- 'using the bysort command'
writeLines(logparse(lf, start = start, end = end))
```

<br>

#### QUICK EXERCISE
> Create a variable for the proportion of the population living in urban
> areas. Use `summarize` to describe your new variable. What's
> the mean and median of your new variable? 

<br>

## Using the `by` and `bysort` commands

Many times we'd like to summarize a variable by subgroups in the
data. For instance, what if we'd like to know which regions have the
highest proportions of children under 5? We could try to use the `by`
command like this, `by region: sum poplt5_pr`, but it won't
work. Stata will refuse to run it because the data are not sorted on the
`region` variable. However, the `bysort` command gives
us an easy way around that problem:

```{r}
start <- 'using the bysort command'
end <- 'univariate graphics: single plot'
writeLines(logparse(lf, start = start, end = end))
```

<br>

#### QUICK EXERCISE
> Create a table of proportion urban by region. Which region has the
> highest proportion of people living in cities? 

<br>  

## Univariate graphics

### `histogram`

To describe a data point, we can use the `histogram` command. If we
want to save the plot, we use the `graph export` command:

```{r}
start <- 'univariate graphics: single plot'
end <- 'univariate graphics with by command'
writeLines(logparse(lf, start = start, end = end))
```

```{r, results = 'asis'}
writeLines(alignfigure('../plots/h_poplt5_pr.png', 'center'))
```

### `histogram` with `by`

You can combine the histogram command with a `by` command to
show the distribution of a variable by groups:

```{r}
start <- 'univariate graphics with by command'
end <- 'kernel density plot'
writeLines(logparse(lf, start = start, end = end))
```

```{r, results = 'asis'}
writeLines(alignfigure('../plots/h_poplt5_pr_reg.png', 'center'))
```

### `kdensity`

You can also use the `kdensity` command to describe the data
using a kernel density plot:

```{r}
start <- 'kernel density plot'
end <- 'generate proportion over 65'
writeLines(logparse(lf, start = start, end = end))
```

```{r, results = 'asis'}
writeLines(alignfigure('../plots/kd_poplt5_pr.png', 'center'))
```

<br>

#### QUICK EXERCISE
> List state name and population less than 5 if population less than 5
> is greater than .1

<br>

## Bivariate graphics

### `scatter`plot

A scatterplot is a very useful tool for the looking at the
relationship between two (or more) variables. Right now I'd like to
look at the relationship between the number of children under 5 and
the number of people over 65. The variable `pop65p` is not a
proportion, so I need to generate a new proportion variable to get
them both on the same scale:

```{r}
start <- 'generate proportion over 65'
end <- 'scatterplot of young population as a function of older population'
writeLines(logparse(lf, start = start, end = end))
```

With my new variable, I can now create a scatterplot: 

```{r}
start <- 'generate proportion over 65'
end <- 'add state labels'
writeLines(logparse(lf, start = start, end = end))
```

```{r, results = 'asis'}
writeLines(alignfigure('../plots/sc_poplt5_pr.png', 'center'))
```

We can add state labels:

```{r}
start <- 'add state labels'
end <- 'change label size'
writeLines(logparse(lf, start = start, end = end))
```

```{r, results = 'asis'}
writeLines(alignfigure('../plots/sc_poplt5_pr_1.png', 'center'))
```

The labels are too big. We can make them smaller.

```{r}
start <- 'change label size'
end <- 'end file'
writeLines(logparse(lf, start = start, end = end))
```

```{r, results = 'asis'}
writeLines(alignfigure('../plots/sc_poplt5_pr_2.png', 'center'))
```

## EXERCISES

1.  Create variables for rate of marriages and divorces
2.  Which region has the highest rates of marriage and divorce in the
    population? 
3.  What do the distributions of these two variables look like?
4.  What does a scatterplot say about the possible relationship
    between the two? 

<br><br>

*Init: 07 June 2015; Updated: `r format(Sys.Date(), format = "%d %B %Y")`*

<br>
