---
title: Programming
author: LPO 9951 | Fall 2015
output: 
  html_document:
   toc: true
   theme: cosmo
   css: ./css/custom.css
---

```{r, echo = FALSE, message = FALSE}
require(knitr)
source('./r/stataknitrhelper.r')
opts_chunk$set(echo = FALSE, message = FALSE, comment = NA, cache = FALSE)
```

```{r, engine = 'bash'}
## run accompanying .do file to get log file for parsing
stata -b -q do ../do/lecture6_programming.do
```

```{r}
## save log file in object
lf <- 'lecture6_programming.log'
```

<br>

#### PURPOSE

Stata programming will save you time, energy, and sanity. Investing
the time now into learning how to program will certainly pay off. It
may seem easy enough now to just copy code 10 times if you need to
complete an operation 10 times, but force yourself to use your
programming skills. By Maymester, you will thank yourself. 

<br>

## Tools you already have

Programming is more than just knowing the most convenient commands to
shorten the time you spend on menial tasks. It involves thinking about
how the commands you do can be combined to make a more efficient,
readable do-file for you and anyone else who will look at it in the
future.

The following points are good places to start when you are trying to
make your program file more efficient.  

* Previous code: You may have already encountered this strategy in the
      work that you have done thus far for the class. Snippets of code
      that you have already toiled over can be used again and
      again. The following tips might come in handy.  
	  * Save your do-files  
	  * Label them well  
	  * Re-use old code, copy-paste  
	  * Make templates if you use a certain piece of code often  
	  * Create files to include or do (e.g., "programs" you can
	  immediately run for things like dealing with missing data)  
* Programming: When you approach your Stata script as a programmer,
  you have a different perspective, a certain general approach on how
  to put these pieces together. The following points are questions you
  might ask yourself in going through the general process for your
  program.  
	  * What is the overall task I am trying to accomplish?  
	  * How are the variables structured? Which variables go together?  
	  * What tasks need to be repeated?  
	  * What procedures may stay the same, though the numerical values
		may change?  




<br>

## Organizing your do file

As your do files increase in length, you will want some type of
organizational structure. A table of contents at the top of the script
can be very helpful. You certainly don't have to do it the way the way
shown below, but you should have something that makes sense to you and
will be clear to others who may read your script.

```{r}
start <- 'h/t to Justin Shepherd'
end <- 'Set preferences/globals/load data'
writeLines(logparse(lf, start = start, end = end))
```

<br>

## File header

Like you've seen in the do files from earlier lectures, it's often
useful to place your file preferences at the top of the script. These
may include, but aren't limited to, graphics settings and global macros
storing directory structurs or url links. If you are only using one
dataset for your analysis, this is a good place to load it.

```{r}
start <- 'Set preferences/globals/load data'
end <- '1\\.0 Describing'
writeLines(logparse(lf, start = start, end = end))
```
<br>

## Describing

`bysort`: Used by itself or in combination with `gen` or `egen`, this
command also allows you to perform a task on numerous categories of a
variable or variables.  

For example, we might want to know what the average flock size is by
status as a loon. We could use the following code:  

```{r}
start <- 'sum with if statement'
end <- 'tab loon with summarize'
writeLines(logparse(lf, start = start, end = end))
```

A slightly easier bit of code would use `tab` with the `summarize` option:

```{r}
start <- 'tab loon with summarize'
end <- 'summarize with bysort'
writeLines(logparse(lf, start = start, end = end))
```

Still another line of code uses the `bysort` command, which takes the
form `bysort <sorting variable>: <command> <variable>`:

```{r}
start <- 'summarize with bysort'
end <- 'summarize multiple variables with bysort'
writeLines(logparse(lf, start = start, end = end))
```

We could actually ask for numerous variables summarized in this way.

```{r}
start <- 'summarize multiple variables with bysort'
end <- '2\\.0 Scalars'
writeLines(logparse(lf, start = start, end = end))
```
<br>

#### Quick Exercise
> Find the average number of feathers in each period by the double
> condition of being a loon and location of nest.

<br>

## Scalars

Scalars temporarily save information that you can use later. There are
two types of information that are stored in STATA after you run
commands. The first is saved as `r` and can be found by using `return
list`. Here are some examples:

```{r}
start <- '2\\.1 return command'
end <- '2\\.2 ereturn command'
writeLines(logparse(lf, start = start, end = end))
```

The second type of information that is stored is under `e`. These can
be found by using `ereturn list`:  

```{r}
start <- '2\\.2 ereturn command'
end <- '2\\.3 scalar command'
writeLines(logparse(lf, start = start, end = end))
```

Keep in mind, however, that each time you run an expression, your
previously stored information in both `return` and `ereturn`
are overwritten. For example, if you run a `sum` command on one
variable, you might have an `r(mean) = 100`. However, the next
time you run `sum` on a new variable, the `r(mean)` will be
overwritten, so you need to be aware of which variable you are using.  

So how do we store this information into memory for future use without
fear of it being overwritten? There are multiple ways to do so. One
easy way includes naming and storing your own scalar using the
`scalar` command, which takes the form of `scalar <name> = <value>`.  

You can name a scalar whatever you want and assign it a value. Let's
do this for the mean of total number of shells in the first period and
show how this preserves the value despite the fact that we run another
mean on feathers.

```{r}
start <- '2\\.3 scalar command'
end <- '3\\.0 Estimates'
writeLines(logparse(lf, start = start, end = end))
```
<br>

#### Quick Exercise
> Use a scalar to calculate the average number of shells across all
> three periods.

<br>

## Estimates

Similar to scalars and returns, estimates store multiple values. This
will be especially useful when we get into regressions next
semester. For now, let's just use estimates to store information we've
learned from the `mean` command.

```{r}
start <- '3\\.1 estimates store command'
end <- '3\\.2 estimates restore command'
writeLines(logparse(lf, start = start, end = end))
```
Now we'll use estimates restore and estimates replay to bring up
previous information that we've stored.

```{r}
start <- '3\\.2 estimates restore command'
end <- '4\\.0 Shortcuts'
writeLines(logparse(lf, start = start, end = end))
```

<br>

## Shortcuts: Numlists and Varlists

Numlists and varlists can make your life much easier by streamlining
your code. Here are some examples of numlists. Notice how we sort the
data using both `sort` and `gsort`. Also notice the `-` sign used in
the second `list` command (`-10/l`) and with `gsort -ideas1`. In the
first case, the sign tells Stata to `list` the last 10 observations
('starting at the end, go back 10). In the second case, Stata
understands that we want to sort our data based on the values in
`ideas1`, but instead of sorting from smallest to largest, as is the
default, we instead want descending values.

```{r}
start <- '4\\.1 numlist commands'
end <- '4\\.2 varlist commands'
writeLines(logparse(lf, start = start, end = end))
```
And here's how we might use varlists. Notice how instead of listing
every variable, we can list the starting and final column with a `-`
between. Using this format requires that we know the order of the
variables in our dataset. We can also use wildcards such as `*`. As
you can see, Stata returns every variable that starts with
`flock`. Keep this feature in mind as you name your variables.

```{r}
start <- '4\\.2 varlist commands'
end <- '5\\.0 Macros'
writeLines(logparse(lf, start = start, end = end))
```
<br>

## Macros

### Globals

We've already been using global macros throughout this course, but it
never hurts to reiterate. Global macros allow you to store many types
of information that will persist throughout a Stata session. We've
been using them to store relative directory links, but they can also
store numerical values and even commands.

Be careful when using global macros. It is easy over the course of a
long Stata session to forget what's hanging around in the memory. To
see which globals (or any macros you have stored for that matter), you
can use the `macro list` command. To drop macros you no longer need, a
generally good policy, use the `macro drop <macro names>` command.

```{r}
start <- '5\\.1 global macros'
end <- '5\\.2 numerical local commands'
writeLines(logparse(lf, start = start, end = end))
```
<br>

### Locals

*Locals* are a way of storing information that you would not really
want to store in a new variable or even scalar. Some of the other automatic results
that are given after running some descriptive or estimation command
are locals. Locals can store a single value or a list of values, but
only for length of time that the script is actively running. This is unlike
global macros, which persist throughout a Stata session (until you
quit the program or purposefully drop them). Once
script has exited, all information stored in locals is lost. There
is a very particular way data in locals are stored and recalled.   

Here are some of the different ways locals are used with numbers:

```{r}
start <- '5\\.2 numerical local commands'
end <- '5\\.3 varlist local commands'
writeLines(logparse(lf, start = start, end = end))
```

Locals can also store strings (such as variable names):

```{r}
start <- '5\\.3 varlist local commands'
end <- '5\\.4 nested local commands'
writeLines(logparse(lf, start = start, end = end))
```

Even better, locals can be nested, that is, a local can hold other
locals:

```{r}
start <- '5\\.4 nested local commands'
end <- '6\\.0 Switch'
writeLines(logparse(lf, start = start, end = end))
```
*NB:* The quotation marks for locals can be tricky. If you are having
trouble getting your locals to do exactly what you want, check to make
sure you are using the correct quotes. The left quote ``` ` ```, or
backtick, is distinct from the normal single quotation mark, `'`. 

<br>

## Switches

Now that weve learned how to create locals, lets use them to create
switches. Switches are impor- tant because you can use them to turn on
and off portions of your code. For example, you can use them to
delegate whether you want to input the full dataset or your most
recently save data. You could also use them to determine whether you
want to turn graphs on or off. They can be very useful.

As an example, below is a switch for turning on or off graphs:

```{r}
start <- 'set switch'
end <- '7\\.0 Loops'
writeLines(logparse(lf, start = start, end = end))
```
As you can see, there is no special switch command. Instead, we create
a local called `graphs` that is set to either 0 or 1. Next comes an
`if` statement that says if the local is equal to 1, then run the
graph. If not, then don't. It's good practice to place your switches
in the top of your do file so that you don't have to hunt for them.  

One trick to keep track of `if` statements (and loops as you will see
below) is to indent code that is within the loop and align the start
of the loop with the end of the loop. It's also good practice with
long loops to use a comment at the closing brace to label the loop
(very helpful when you have many loops in your file).

<br>

## Loops

Loops are used when you are performing one task on a variable or group
of variables and `bysort` and `egen` cannot meet your needs. It is useful
to think about how the procedures you are running can be grouped
together and how the same structure can be applied to multiple cases.
Stata loops have a few different structures:  

  * `if` (`if` / `else`)
  * `foreach`
  * `forvalues`
  * `while`

<br>

### `if` / `else`

To start, let's just use our switch structure to specify an
alternative action if the switch condition is not met.

```{r}
start <- '7\\.1 if/else command'
end <- '7\\.2 foreach command'
writeLines(logparse(lf, start = start, end = end))
```

If `switch == 1` then the first command will run; in all other cases,
the second command will run. 

<br>

### `foreach`

Another type of loop uses the `foreach` command. Take a look at the
[help file](http://www.stata.com/manuals14/pforeach.pdf) for `foreach`
statements. As you can see, there are a variety 
of different ways to use the `foreach` command. Here are some
examples:

```{r}
start <- '7\\.2 foreach command'
end <- '7\\.3 forvalues command'
writeLines(logparse(lf, start = start, end = end))
```
<br>

#### Quick Exercise
> Rescale each `shells*` variable so it is in 1000s of shells.

<br>

### `forvalues`

Another loop command that is quite useful is called `forvalues`. The
`forvalues` loop uses a counter within a loop and repeats the loop
until you hit the maximum specified value. Here are some examples;
notice the different ways to count:  

```{r}
start <- '7\\.3 forvalues command'
end <- '7\\.4 while command'
writeLines(logparse(lf, start = start, end = end))
```
<br>

#### Quick Exercise
> Use `forvalues` to create means for days in nest.

<br>

### `while`

Finally, `while` loops are another way to loop using numbers. They are
similar to `forvalue`s loops in Stata, but require a counter. Though
the two are generally interchangeable, while loops are technically
about waiting to fulfill a condition. Therefore, they can be used in
more ways than `forvalues` loops. Keep in mind, however, that if you
set a condition that will never be fulfilled, your `while` loop will
run forever (or until your computer crashes or the network
administrator, if you are running code through a network, kills the
process and sends you a mean email).

```{r}
start <- '7\\.4 while command'
end <- '8\\.0 Nests'
writeLines(logparse(lf, start = start, end = end))
```

<br>

## Nests

It is also possible to nest loops within loops. When you do this, the
outer loop runs until it hits an inner loop. Then it evaluates the
inner loop until the inner loop is finished. Then it will 
continue with the outer loop. If the inner loops statement uses an
`if` statement, Stata will only evaluate it if the condition is met
(evaluates to true). This can get very
complicated very quickly, so you need to know where you are in the
code. This is why it is smart to indent all commands within a loop to
the level of the loop.

```{r}
start <- 'set up locals for nest'
end <- 'end file'
writeLines(logparse(lf, start = start, end = end))
```

<br>

#### Quick Exercise
> Using the following local set up, create a nested loop that uses
> either `forvalues` or a `while` loop to summarize all the listed
> variables by period, conditional on the status as a loon.
> 
> ```
> local all1 shells1 feathers1 flock1 ideas1 eggs1 nest1
> local all2 shells2 feathers2 flock2 ideas2 eggs2 nest2
> local all3 shells3 feathers3 flock3 ideas3 eggs3 nest3
> ```
<br>

## Sectioning your do-file (templates)

You will go through the same general procedures every time you work
with quantitative data. The structure of this class is a good guide
for you to create your own template do-file that you can pull up every
time you start a new research project. Sections might include the
following:

  * Setting up Stata (most of what the do files we have been using for
  class already have)
  * Setting up globals/locals/file preferences
  * Pulling in the data you will use
  * Data cleaning/validation
  * Taking account of the survey design
  * Descriptive statistics
  * Regression model(s)
  * Recording output

<br>
<br>

*Init: 16 August 2015; Updated: `r format(Sys.Date(), format = "%d %B %Y")`*

<br>
